Filamentary transport in high-power H-mode conditions and in no/small-ELM regimes to predict heat and particle loads on PFCs for future devices

### 2016-11-11 Presentation/gpm_november2016_vianello.tex
Further modifications
Nicola please take a look

	modified:   gpm_november2016_vianello.tex

\section{Background}
\begin{frame}{Present status}
  \setbeamercovered{transparent}
  \begin{enumerate}[<+(1) | invisible@-+>]
    \item Experimental and theoretical investigations suggest that SOL density profiles are  determined by the balance between radial filamentary transport and parallel dynamics
    \item The balance changes at high density and leads to broadened SOL density profiles $\rightarrow$ \emph{shoulder formation}
    \item This is observed in L-mode plasmas in a variety of devices including AUG \parencite{Carralero:2014gs},
      TCV \parencite{Garcia:2007p2615} and MAST \parencite{Militello:2016hk}
    \item Observations exhibit various dependencies on divertor
      collisionality, plasma current,  seeding. \alert{Results from
        different devices not fully reconciled and the underlying mechanism is not fully understood.}
    \item Studies of shoulder formation in H-Mode are so far limited 
	%\item Particle and energy transport in the SOL is crucial for lifetime of PFC in ITER and DEMO
%    \item First we will report a summary of the achievements and then
 %    personal comments and plans
    \end{enumerate}
\end{frame}
\section{L-Mode}
\begin{frame}{L-Mode studies: AUG/1}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \begin{itemize}
      \item<1-> AUG and JET \parencite{Carralero:2015gu} suggest that
        \begin{equation*}
        \Lambda_{div} = \frac{L_{\parallel}/c_s}{1/\nu_{ei}}\frac{\Omega_i}{\Omega_e}
      \end{equation*}
      controls shoulder formation
        \item<2-> \textcolor{red}{Tested by changing n$_e$
            and $T_e$ through fueling/seeding/heating}
        \item<3-> This determines a change of the velocity-size
          scaling from
        \textcolor{ta3chameleon}{sheath-limited}
        to \textcolor{blue}{inertial regime}. \onslide<4>{\alert{$\Lambda_{div}$ rules the density profile scale
          length}} 
        \end{itemize}
    \end{column}
    \begin{column}{0.5\textwidth}
      \only<1-2>{
      \begin{tikzonimage}[width=\textwidth]{../pdfbox/KoM15Nov/CarraleroPRL15a}
          \draw [thick, ta3chameleon, thick, rotate=1] (0.43,0.25) ellipse (0.23 and 0.1);
          \draw [thick, blue, thick, rotate=-4] (0.8,0.55) ellipse (0.1 and 0.33);
        \only<2>{
          \draw [ultra thick, red, thick, dashed] (0.5,0.63) circle (0.12);
        }
      \end{tikzonimage}}
    \only<3>{
      \begin{tikzonimage}[width=\textwidth]{../pdfbox/KoM15Nov/CarraleroPRL15b}
          \draw [ultra thick, ta3chameleon, rotate=1] (0.2,0.5) ellipse (0.05 and 0.4);
          \draw [thick, blue, rotate=15] (0.65,0.07) ellipse (0.4 and 0.1);
      \end{tikzonimage}
    }
    \includegraphics<4>[width=\textwidth]{../pdfbox/KoM15Nov/CarraleroPRL15c}
    \end{column}
  \end{columns}
\end{frame}
\begin{frame}{L-Mode studies: AUG/2}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \begin{itemize}
      \item<1-> Profile modified by increase of blob-size and change of
        \alert{packing fraction: $f_{fil} = \nu_{fil}\tau_{AC}$}
        \onslide<2->{\textcolor{blue}{and filament relative density}
          {\footnotesize (Carralero 2016 in preparation)}}
      \item<3-> As a consequence the contribution of filaments to
        radial transport increases
      \item<4> Parallel flow is strongly reduced whenever we increase
        the divertor collisionality
      \end{itemize}
    \end{column}
      \begin{column}{0.5\textwidth}
        \includegraphics<1>[width=\textwidth]{../pdfbox/KoM15Nov/CarraleroMST16a}
        \includegraphics<2>[width=\textwidth]{../pdfbox/KoM15Nov/CarraleroMST16b}
        \includegraphics<3>[width=\textwidth]{../pdfbox/KoM15Nov/CarraleroMST16c}
        \includegraphics<4>[width=\textwidth]{../pdfbox/KoM15Nov/CarraleroMST16d}
      \end{column}
    \end{columns}
  \end{frame}

@@

  \section{Open issues}
  \begin{frame}{Open and unresolved issues}
  \setbeamercovered{transparent}
  \begin{enumerate}[<+(1) | invisible@-+>]
    \item Does $\Lambda_{div}$ unify the picture among the devices?
      \alert{No, a dependence on neutral pressure is likely to play a role}
    \item If neutrals play a role is it at the divertor and/or at the midplane?
      \alert{Contradictory results if one include experiments with
        midplane puffing, JET and EIRENE simulations}
    \item Does L$_\parallel$ play any role? \alert{So far TCV suggests no
      dependence from f$_x$ scan but both TCV and MAST observe an
      I$_p$ dependence. MAST clearly state filament $\sigma_{\perp}$
      increases with L$_{\parallel}$ }
    \item Is cooling the divertor with fueling or with equivalent? \alert{Contradictory observations in AUG and JET}
    \item Do we observe same behavior in L and H-Mode? \alert{So
      far no as shown in H-Mode AUG. We need higher detachment
      condition and we need enough fueling}
      \item What are the underlying mechanisms? 
    \end{enumerate}
  \end{frame}

\section{Topic 21 experiments }
\begin{frame}{$n_{\text{proposed shots}} \gg n_{\text{allocated shots}}$}
	\begin{itemize}
		\item 15 proposals submitted to Topic 21
		\item Proposals include experiments on all three machines 
		\item There are overlaps between several of the proposals 
		\item Preliminary shot allocation. AUG: 14. MAST: 13. TCV: 23 !!!!!!!1SHOULD WE ADD PRELIMINARY DATES FOR THE EXPERIMENTS??
		\item However, total number of proposed shots:  \textbf{449}
		\item Several of the proposed experiments can be combined. \underline{But we must prioritize}
		\item Reaching all goals is not possible with the current number of allocated shots.
	\end{itemize}
\end{frame}

\begin{frame}{Focal points}
\begin{itemize}
	\item MST1 uniquely facilitates cross comparison between machines
	\item Several of the proposed experiments overlap 
	\item A cross machine experiment has the makings of settling open issues
	\item Therefore, we will allocate shots for cross-machine comparison
	\item L-mode experiments: 
	\begin{enumerate}
		\item Investigate the role of neutrals 
		\item $I_p$ and $q_{95}$ scans
	\end{enumerate}
	\item H-mode experiments. 
\end{itemize}	
\end{frame}

@@

\begin{frame}{H-mode experiments}
\begin{itemize}
	\item ITER and DEMO will operate in H-mode
	\item We must know what parameters control shoulder formation 
	\item Shoulder formation parameter regime is unclear on all machines
	\item Main priorities: 
	\begin{enumerate}
		\item Investigate if clear shoulder formation exists and what plasma parameters required? 
		\item Investigate the SOL (filamentary) transport properties. Main diagnostic is reciprocating probes $\rightarrow$ limits heating power
		\item Experiments must gradually increase power and density. 
	\end{enumerate}
	\item Fueling in H-mode is problematic due to transport barrier? NBI fueling?
	\item H-mode density limit\parencite{Bernert:2015ppcf} must be avoided
\end{itemize}

@@

\begin{frame}{Optimizing cross machine comparison}
\begin{itemize}
	\item Machines are fundamentally differently designed
	\item Strive for similar configurations of the machines:
	\begin{enumerate} 
		\item Single-Null
		\item $\bm{B} \times \nabla B$ towards active divertor 
		\item Strike-point and cryo pump location (AUG and MAST-U)
		\item Heating 
	\end{enumerate}
\end{itemize} 
%things we cannot influence but could be important : wall (carbon, tungsten), helicity, aspect ratio, divertor design,

@@

\begin{frame}{Proposed shot plan - AUG}
\vspace{-1cm}
Allocated shots $\sim$ 14 + contingency 
\begin{itemize}
	\item L-Mode (6 shots)
	\begin{enumerate}
		\item Reference shot \# 30276
		\item Perform $I_p$ scan (three values) with fixed $B_T$
		\item Perform $I_p$ scan fixing  $q_{95}$	 ($B_T$)	
		\item $B_T= 2.0$ required by proposal 6 (Aguiam). Should be possible for some shots? 		
		\item should we add OM puff to validate neutral measurements?
		\item These shots combines experiments proposed by: Carralero, Militello, 	Vianello

@@

\begin{frame}{Preliminary shot plan  - MAST-U}
	\begin{itemize}
		\item 
	\end{itemize}
\end{frame}

\begin{frame}{There are urgent experiment which cannot be addressed}
\begin{itemize}
	\item Ion temp measurements
	\item Topology investigations
	\item Strikepoint sweeping and cryo pump
	\item Parameters scans in H-mode
	\item Strike-point sweep

\end{itemize}
\end{frame}

\begin{frame}{Future and Miscellaneous}
	\begin{itemize}
		\item piggybag on Quescient H-mode shots
		\item Shortly after this meeting we will organize a follow-up meeting to reconcile ideas from this discussion (date not fixed yet)
		\item All information will be gathered on the wiki

		\end{itemize}
\end{frame}

>>>>

\begin{frame}{Outline}
	\begin{itemize}
		\item Background 
		\item Plans
		\item Discussion
	\end{itemize}	
\end{frame}

\section{Background}
\begin{frame}{Present status}
  \setbeamercovered{transparent}
  \begin{enumerate}[<+(1) | invisible@-+>]
    \item Experimental and theoretical investigations suggest that SOL density profiles are  determined by the balance between radial filamentary transport and parallel dynamics
    \item The balance changes at high density and leads to broadened SOL density profiles $\rightarrow$ \emph{shoulder formation}
    \item This is observed in L-mode plasmas in a variety of devices including AUG \parencite{Carralero:2014gs},
      TCV \parencite{Garcia:2007p2615} and MAST \parencite{Militello:2016hk}
    \item Observations exhibit various other dependencies e.g., divertor
      collisionality, plasma current, seeding. \alert{Results from
        different devices not fully reconciled and the underlying mechanism is not fully understood.}
    \item Studies of shoulder formation in H-Mode are so far limited 
	%\item Particle and energy transport in the SOL is crucial for lifetime of PFC in ITER and DEMO
%    \item First we will report a summary of the achievements and then
 %    personal comments and plans
    \end{enumerate}
\end{frame}
\section{L-Mode}
\begin{frame}{L-Mode studies: AUG/1}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \begin{itemize}
      \item<1-> AUG and JET \parencite{Carralero:2015gu} suggest that divertor collisionality 
        \begin{equation*}
        \Lambda_{div} = \frac{L_{\parallel}/c_s}{1/\nu_{ei}}\frac{\Omega_i}{\Omega_e}
      \end{equation*}
      controls shoulder formation
        \item<2-> ... which is associated with a change of the velocity-size
          scaling from
            \textcolor{ta3chameleon}{sheath-limited}
        to \textcolor{blue}{inertial regime}.
        \item<3-> \textcolor{red}{Tested by changing n$_e$
            and $T_e$ through fueling/seeding/heating}

       \onslide<4>{\alert{$\Lambda_{div}$ rules the density profile scale
          length}} 
        \end{itemize}
    \end{column}
    \begin{column}{0.5\textwidth}
      \only<1-2>{
      \begin{tikzonimage}[width=\textwidth]{../pdfbox/KoM15Nov/CarraleroPRL15a}
          \draw [thick, ta3chameleon, thick, rotate=1] (0.43,0.25) ellipse (0.23 and 0.1);
          \draw [thick, blue, thick, rotate=-4] (0.8,0.55) ellipse (0.1 and 0.33);
        \only<2>{
          \draw [ultra thick, red, thick, dashed] (0.5,0.63) circle (0.12);
        }
      \end{tikzonimage}}
    \only<3>{
      \begin{tikzonimage}[width=\textwidth]{../pdfbox/KoM15Nov/CarraleroPRL15b}
          \draw [ultra thick, ta3chameleon, rotate=1] (0.2,0.5) ellipse (0.05 and 0.4);
          \draw [thick, blue, rotate=15] (0.65,0.07) ellipse (0.4 and 0.1);
      \end{tikzonimage}
    }
    \includegraphics<4>[width=\textwidth]{../pdfbox/KoM15Nov/CarraleroPRL15c}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{L-Mode studies: AUG/2}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \begin{itemize}
      \item<1-> Profile modified by an increase of blob-size and change of
        \alert{packing fraction: $f_{fil} = \nu_{fil}\tau_{AC}$}
        \onslide<2->{\textcolor{blue}{and filament relative density}
          {\footnotesize (Carralero 2016 in preparation)}}
      \item<3-> As a consequence the contribution of filaments to
        radial transport increases
      \item<4> Parallel flow is strongly reduced whenever we increase
        the divertor collisionality
      \end{itemize}
    \end{column}
      \begin{column}{0.5\textwidth}
        \includegraphics<1>[width=\textwidth]{../pdfbox/KoM15Nov/CarraleroMST16a}
        \includegraphics<2>[width=\textwidth]{../pdfbox/KoM15Nov/CarraleroMST16b}
        \includegraphics<3>[width=\textwidth]{../pdfbox/KoM15Nov/CarraleroMST16c}
        \includegraphics<4>[width=\textwidth]{../pdfbox/KoM15Nov/CarraleroMST16d}
      \end{column}
    \end{columns}
  \end{frame}

@@

  \section{Open issues}
  \begin{frame}{Open and unresolved issues}
  \setbeamercovered{transparent}
  \begin{enumerate}[<+(1) | invisible@-+>]
    \item Does $\Lambda_{div}$ unify the picture among the devices?
      \alert{No, a dependence on neutral pressure is likely to play a role}
    \item If neutrals play a role is it at the divertor and/or at the midplane?
      \alert{Contradictory results if one include experiments with
        midplane puffing, JET and EIRENE simulations}
    \item Does L$_\parallel$ play any role? \alert{So far TCV suggests no
      dependence from f$_x$ scan but both TCV and MAST observe an
      I$_p$ dependence. MAST clearly state filament $\sigma_{\perp}$
      increases with L$_{\parallel}$ }
    \item Is cooling the divertor with fueling or with equivalent? \alert{Contradictory observations in AUG and JET}
    \item Do we observe same behavior in L and H-Mode? \alert{So
      far no as shown in H-Mode AUG. We need higher detachment
      condition and we need enough fueling}
      \item Is the density shoulder accompanied with an ion temperature shoulder?
%      \item What are the underlying mechanisms? 
    \end{enumerate}
  \end{frame}

\section{Topic 21 experiments }
\begin{frame}{$n_{\text{proposed shots}} \gg n_{\text{allocated shots}}$}
	\begin{itemize}
		\item 15 were proposals submitted to Topic 21
		\item Proposals include experiments on all three machines 
		\item There are overlaps between several of the proposals 
		\item Preliminary shot allocation. AUG: 14. MAST: 13. TCV: 23 !!!!!!!1SHOULD WE ADD PRELIMINARY DATES FOR THE EXPERIMENTS??
		\item However, total number of proposed shots:  \textbf{449}
		\item Several of the proposed experiments can be combined. \underline{But we must prioritize}
		\item Reaching all goals is not possible with the current number of allocated shots.
	\end{itemize}
\end{frame}

\begin{frame}{Focal points}
\begin{itemize}
	\item MST1 uniquely facilitates cross comparison between machines
	\item Several of the proposed experiments overlap 
	\item A cross machine experiment has the makings of settling open issues
	\item Therefore, we will allocate shots for cross-machine comparison
	\item ....but also to other proposed experiments
	\item Cross machine L-mode experiments: 
	\begin{enumerate}
		\item Investigate the role of neutrals 
		\item $I_p$ and $q_{95}$ scans
	\end{enumerate}
	\item Cross machine H-mode experiments. 
\end{itemize}	
\end{frame}

@@

\begin{frame}{H-mode experiments}
\begin{itemize}
	\item ITER and DEMO will operate in H-mode
	\item We must know what parameters control shoulder formation 
	\item Shoulder formation parameter regime is unclear on all machines
	\item Main priorities: 
	\begin{enumerate}
		\item Investigate if clear shoulder formation exists and what plasma parameters required? 
		\item Investigate the SOL (filamentary) transport properties. Main diagnostic is reciprocating probes $\rightarrow$ limits heating power
		\item Experiments must gradually increase power and density. 
	\end{enumerate}
	\item Fueling in H-mode is problematic due to transport barrier? NBI, pellet fueling?
	\item H-mode density limit\parencite{Bernert:2015ppcf} must be dealt with
\end{itemize}

@@

\begin{frame}{Optimizing cross machine comparison}
\begin{itemize}
	\item Machines are fundamentally differently designed
	\item Strive for similar configurations of the machines:
	\begin{enumerate} 
		\item Single-Null
		\item $\bm{B} \times \nabla B$ towards active divertor 
		\item Strike-point and cryo pump location (AUG and MAST-U)
		\item Heating 
		\item Fueling location 
		\item Seeding (species, location, rate)
	\end{enumerate}
\end{itemize} 
%things we cannot influence but could be important : wall (carbon, tungsten), helicity, aspect ratio, divertor design,

@@

\begin{frame}{Proposed shot plan - AUG}
\vspace{-1cm}
Allocated shots $\sim$ 14 + contingency 
\begin{itemize}
	\item L-Mode (6 shots)
	\begin{enumerate}
		\item Reference shot \# 30276
		\item Perform $I_p$ scan (three values) with fixed $B_T$
		\item Perform $I_p$ scan fixing  $q_{95}$	 ($B_T$)	
		\item $B_T= 2.0$ required by proposal 6 (Aguiam). Should be possible for some shots? 		
		\item should we add OM puff to validate neutral measurements?
		\item Possibility of strike-point sweeping during shot
		\item RFEA measurements in internal program. 
		\item These shots combines experiments proposed by: Carralero, Militello, 	Vianello

@@

\begin{frame}{Preliminary shot plan  - MAST-U}
	\begin{itemize}
		\item L-mode (6 shots)
		\begin{enumerate}		
			\item New machine. No reference shot yet			
			\item Same $I_p$ and $q_{95}$ scans as on AUG. Availability of multi pin probe? 
			\item Try varying fueling location
			\item RFEA measurements during shoulder formation 
		\end{enumerate}
		\item H-mode (7 shots)
		\begin{enumerate}
			\item These experiments require the existence of H-mode reference shot. Perhaps not available in 2017
			\item Similar scenario development as on other machines
		\end{enumerate}

					\item Same $I_p$ and $q_{95}$ scans as on AUG
	\end{itemize}
\end{frame}

\begin{frame}{There are urgent experiment which cannot be addressed}
\begin{itemize}
	\item Detailed ion temp measurements 
	\item Topology investigations
	\item Strikepoint sweeping and cryo pump
	\item Parameters scans in H-mode
	\item X-point probes
\end{itemize}
\end{frame}

\begin{frame}{Future and Miscellaneous}
	\begin{itemize}
		\item Possibility to piggybag on Quescient/small ELM H-mode shots (Topic 5,6, 18)
		\item Shortly after this meeting we will organize a follow-up meeting to 						reconcile ideas from this discussion (date not fixed yet)
		\item All information will be gathered on the wiki
		\item Filament database		
	\end{itemize}
\end{frame}
